*Specific taxa differentiate proximal and distal

### 2017-06-18 submission/manuscript.Rmd
text updated with Hannigan edits

---
title: '**Machine learning identifies unique taxa differentiating proximal and distal
  human colonic microbiota*'
csl: mbio.csl
fontsize: 11pt
output:
  html_document: default
  pdf_document:
    includes: null
    keep_tex: yes
  word_document: default
geometry: margin=1.0in
bibliography: references.bib
---
\vspace{35mm}

@@

Colorectal cancer (CRC) remains a leading cause of death worldwide. Tumors of the proximal (right) and distal (left) colon are morphologically and genetically distinct. Previous work from our group found that microbial dysbiosis is associated with the development of CRC tumors in studies of both mice and humans. Analysis of the fecal microbiota from healthy and CRC patients further revealed different microbial signatures associated with disease. In this study, we extended our observations of the fecal microbiome to analysis of the proximal and distal human colon. We used a two-colonoscope approach on subjects that had not undergone standard bowel preparation procedure. This technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption. 16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and exit stool for 20 healthy individuals. Diversity analysis revealed that each site contained a diverse community, and that a patient’s samples were more similar to each other than to that of other individuals. Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the machine-learning algorithm Random Forest to identify key species that distinguish biogeographical sites. Random Forest classification models were built using taxa abundance and sample location and revealed distinct populations that were found in each location. _Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas_ and _Actinomyces_ were most likely to be found in mucosal samples versus luminal samples (AUC = 0.925). The classification model performed well (AUC = 0.912) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = 0.755). The left mucosa was found to have high populations of _Finegoldia, Murdochiella_ and _Porphyromonas_. Proximal and distal luminal samples were comprised of many of the same taxa, likely reflecting the fact that stool moves along the colon from the proximal to distal end. Finally, comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the left lumen, again reflecting the anatomical structure of the colon. Taken together, our results have identified distinct bacterial populations of the proximal and distal colon. Further investigation of these bacteria may elucidate if and how these groups contribute to differential oncogenesis processes on the respective sides of the colon.

@@

Colorectal cancer (CRC) is the second-leading cause of cancer-related deaths in the United States. CRC tumors vary in structure, size, morbidity and symptomology depending upon their geographical location in the colon. CRC has recently been described to fall into four molecular subtypes (@Guinney2015). Of these subtypes, tumors that arise on the left side of the colon are infiltrating lesions that present with painful symptoms. In contrast, right-sided colon tumors are bulky and project into the lumen, often causing no symptoms until they have developed into carcinomas (@Benedix2010). Because of the absence of symptoms, right-sided tumors, which comprise up to 47% of CRCs, have a significantly higher mortality rate (@Benedix2010). The left and right sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, as well as oxygen, pH and the presence of antimicrobial peptides (@Yamauchi2012, @Albenberg2014, @Donaldson2015). Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. Given this varied physiology of the proximal-distal axis of the colon, symbiotic microbes and their metabolites likely vary between sites.

Several recent findings have shown that development and progression of CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012). Our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (@Zackular2014). Further studies manipulating the gut microbiome using antibiotics or other chemoagents in mice has shown that dysbiosis preceded and accelerated the development of CRC tumors (@Zackular2013). Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated based on FISH staining and 16S rRNA sequencing (@Dejea2014). These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_. Interestingly, the periodontal pathogens are also highly predictive of if a patient has CRC tumors or not in our classification studies (@Baxter2016).  However, these studies have been performed comparing shed stool or in mice, which does not allow for analysis of paried samples from the proximal and distal sides of the colon. Similarly, comparisons of on- or off-tumor bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site. Disruption of the healthy community could promote the initiation or proliferation of the distinct left and right CRC tumors. However, the contribution of the gut microbiota to these subtypes is largely undefined.

Furthermore, profiles of the microbial biogeography of the gut have been limited by sample collection methods. The majority of human gut microbiome studies have been performed on whole shed stool or on samples collected during colonoscopy procedures. While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel. Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer (@Jalanka2014, @Harrell2012). As such, what little information we do have about the biogeographical distribution of the microbes in the proximal and distal colon is confounded by the bowel preparation procedure.

In this study we sought to combine these approaches to characterize the microbiome in the lumen and mucosa of the proximal and distal healthy human colon. Our design uses unprepared colonoscopy techniques to sample each location of the gut without prior disruption of the native flora in 20 healthy volunteers. To get around the noise created by a diverse set of human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location. We found that our classification models were able to separate mucosal and lumenal samples as well as differentiate between sides of the colon based on populations of specific microbes. By identifying the specific microbes we are poised to ask if and how the presence or disruption of the microbes at each site contribute to the development of the specific tumor subtypes of CRC in the proximal and distal human colon.

@@

#### Microbial membership and diversity of the proximal and distal colon 
Proximal and distal lumenal and mucosal samples were collected from 20 healthy humans that had not undergone bowel preparation(Figure 1). Participants also collected stool at home one week prior to the procedure. To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample. The relative abundance of each phyla for each site is shown in Figure 2A. This data combines the samples from all participants in the study. Each site is primarily dominated by _Firmicutes_ and _Bacteriodetes_, consistent with known variability in human microbiome research. Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B). That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa, and so on. Therefore we could not identify a clear pattern of changes in microbial diversity along the gut axis.

To compare similarity between sides (left or right) or sites (lumen or mucosa), we calculated theta YC distances from OTU abundances and compared distances for all individuals. Again, across all patient samples we saw a range of theta YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis. However, when comparing median distances between the right lumen and mucosa, the left versus right lumen, the left versus right mucosa and the left lumen and mucosa, we found that the right lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment). Next, we calculated thetaYC distances to examine how each sample compared to the home-collected exit stool. Amidst variability between patients, we did identify significantly smaller thetaYc distance between the left lumenal sample and the exit stool (Figure 3B, P < 0.05, Wilcoxon, BH adjustment). Further, there was an even larger difference in the comparisons of the left mucosa to the exit stool, indicating that the mucosa is quite different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment). To determine what factors may be driving the differences seen among the samples, we compared thetaYC distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal). We found that samples from one individual were overall much more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon (@Eckberg2005). Thus interpersonal variation between subjects drives the differences between samples more than sample site or location. Overall, the results of the similarity analysis suggest that the contents of the left lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different.

#### Random Forest classification models identify important OTUs on each side
To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest machine-learning models built on OTU abundances. The first model we built to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A). The constructed model used ((Xopt)) features for the right and ((Xopt)) for the left. The models performed well when classifying these samples (AUC = 0.797 and AUC = 0.980, respectively). The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the right lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B). The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4C). These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.

@@

#### Bacterial OTUs associated with cancer are found in healthy individuals 
Given that specific bacterial species have been associated with colorectal cancer we probed our sample set for these OTUs. Among our 100 samples, the most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subsp animalis_. It was the only sequence to align with a _F. nucleatum_ strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors. (@Castellarin2011). The _Fusobacterium_ positive samples were located on the right and left mucosa and ranged up to 1% of the sample (Figure 6A). OTU00152 is _Porphyromonas_ and the most frequent sequence in that OTU aligns to _Porphyromonas asacharolytica_, another bacterium commonly detected and isolated from colorectal tumors. OTU152 was only detected on the left mucosa, and in fact was one of the OTUs the classification model identified as separating left and right sides (Figure 6B). Samples that were positive for _P. asacharolytica_ ranged in abundances from 0.01% - 16%. Thus, cancer-associated OTUs could be found in our sample set of 20 healthy individuals.

@@

When examining the relative abundance of the different phyla at each site, there is a wide amount of variation for each phyla with communities primarily dominated by the Bacteriodes and Firmicutes. This likely reflects not only the variability between human subjects, casued by differences in age, gender, diet, but also reports of microbial "patchiness" in the gut microbiome. Several previous studies have noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other (@Hong2011). Similar patchiness is also observed in lumenal contents and fecal samples themselves; there is observed separation of different interacting microbes along the length of a stool sample, for instance (@Stearns2011). That said, across our samples the mucosal samples harbor more Proteobacteria, consistent with previous studies comparing mucosal swabs to lumenal content in humans (@Albenberg2014). Hence, the conclusions we can draw from phyla analysis are likely confounded by differences in sampling and patchiness between subjects.

We identified a range of Simpson diversity values between patients and between each anatomical site. As this measurement takes both number of species and abundance of species into account, it seems that there is a varied level of each component along the human colon. There was no clear pattern of diversity when comparing patients- some patients had a higher level of diversity in the proximal than the distal colon, some patients had the opposite. An early study of microbial diversity and membership of three human subjects found a varied level of diversity in each subject, but that subjects were overall more similar to themselves than to each other (@Eckberg2005). We used the thetaYC dissimilarity metric to compare the communities present at each anatomical location to each other. We found that while there was a wide range of similarity of each site within a person, the study participants were ultimately more similar to themselves than all of the other participants. This finding is consistent with numerous studies examining the biogeography of the colon in healthy humans (@Eckberg2005, @deCarcer2010, @Zhang2013).

@@

We detected _F. nucleatum_ and _P. asacharolytica_ in 7 and 5 of our subjects, respectively. These bacteria have been shown to be predictive of colorectal cancer in humans (@Baxter2016) and have oncogenic properties in cell culture and in mice (@Sears2014). Interestingly, while _F. nucleatum_ was found on both sides of the colon, _P. asacharolytica_ was only detected in the left mucosa. Not much is known about the distribution of _P. asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous left mucosa (@Albenberg2014). In studies examining bacteria on colorectal cancer tumors, _F. nucleatum_ is more commonly detected on right-sided tumors, and distribution of _F. nucleatum_ decreases along the colon to rectum (@Mima2016). Of the (8) (40%) individuals positive for _F. nucleatum_ in this study, the bacteria was spread across the right mucosa, left lumen and left mucosa. That said, in a previous study of 1,102 colorectal cancer cases, only 138 (11%) of them contained detectable _F. nucleatum_ at all (@Mima2016). In a separate study of patients with inflammatory bowel disease (IBD), 50% of 22 IBD patients and 17.6% of 34 people without disease contained _F. nucleatum_ in mucosal biopsies (@Strauss2011). These studies have the added caveat of being sampled from prepared or cleaned colons. Thus, the presence of _F. nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers. Because of the spatial distribution of the _F. nucleatum_ in our sample set, we cannot develop a model for the role of _F. nucleatum_ in the healthy colon. Indeed, rather than the _F. nucleatum_ initiating tumorigenesis, these bacteria may instead arrive "late" to the tumor microenvironment and feed on tumors that have alread begun to develop (@Flynn2016). Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found both on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis. Finally, _Fusobacteria_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model (@Whitmore2014, @Flynn2016). Thus, further analysis of the distribution and role of these pathogens may elucidate a mechanism for development of CRC tumors in the proximal or distal colon.

>>>>

---
title: '*Specific taxa differentiate proximal and distal
  microbiota in the unprepped human colon*'
csl: mbio.csl
fontsize: 11pt
output:
  html_document: default
  pdf_document:
    includes: null
    keep_tex: yes
  word_document: default
geometry: margin=1.0in
bibliography: references.bib
---
\vspace{35mm}

@@

Colorectal cancer (CRC) remains a leading cause of death worldwide despite improved preventative and therapeutic measures. Tumors of the proximal (right) and distal (left) colon are morphologically and genetically distinct. Previous work from our group found that microbial dysbiosis is associated with the development of CRC tumors in studies of both mice and humans. Analysis of the fecal microbiota from healthy and CRC patients further revealed different microbial signatures associated with disease. We extended our observations of the fecal microbiome to include analysis of the proximal and distal human colon. We used a two-colonoscope approach on subjects that had not undergone standard bowel preparation procedure. This technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption. 16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and home-collected stool for 20 healthy individuals. Diversity analysis revealed that each site contained a diverse community, and that a patient’s samples were more similar to each other than to that of other individuals. Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the Random Forest machine-learning algorithm to identify key species that distinguish biogeographical sites. Random Forest classification models were built using taxa abundance and sample location and revealed distinct populations that were found in each location. _Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas_ and _Actinomyces_ were most likely to be found in mucosal samples versus luminal samples (AUC = 0.925). The classification model performed well (AUC = 0.912) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = 0.755). The left mucosa was found to have high populations of _Finegoldia, Murdochiella_ and _Porphyromonas_. Proximal and distal luminal samples were comprised of many of the same taxa, likely reflecting the fact that stool moves along the colon from the proximal to distal end. Finally, comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the left lumen, again reflecting the anatomical structure of the colon. By sampling the unprepped human colon, our results have identified distinct bacterial populations native to the proximal and distal sides. Further investigation of these bacteria may elucidate if and how these groups contribute to differential oncogenesis processes on the respective sides of the colon.

@@

Colorectal cancer (CRC) is the second-leading cause of cancer-related deaths in the United States. CRC tumors vary in structure, size, morbidity and symptomology depending upon their geographical location in the colon. CRC has recently been described to fall into four molecular subtypes (@Guinney2015). Of these subtypes, tumors that arise on the left side of the colon are infiltrating lesions that present with painful symptoms. In contrast, 47% of CRCs are caused by right-sided colon tumors that are bulky and project into the lumen, often remaining asymptomatic until advanced carcinogenesis (@Benedix2010). Due to the absence of symptoms, right-sided tumors have a significantly higher mortality rate (@Benedix2010). The left and right sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, as well as oxygen, pH and the presence of antimicrobial peptides (@Yamauchi2012, @Albenberg2014, @Donaldson2015). Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. Given this varied physiology of the proximal-distal axis of the colon, symbiotic microbes and their metabolites likely vary between sites.

Several recent findings have shown that development and progression of CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012). Our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (@Zackular2014). Further studies manipulating the gut microbiome using antibiotics or other chemoagents in mice has shown that dysbiosis preceded and accelerated the development of CRC tumors (@Zackular2013). Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated (@Dejea2014). These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_. Interestingly, these periodontal pathogens have been highly predictive of whether a patient had CRC tumors or not in our classification studies (@Baxter2016).  Because these studies were performed in mice or examined only shed human stool, they were unable to analyze paried samples from the proximal and distal sides of the colon. Similarly, comparisons of on- or off-tumor bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site. Due to these limitations, the contribution of the gut microbiota to these subtypes is largely undefined. Characterizing these communities could provide needed insight into CRC etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct left and right CRC tumors.

Further, the few existing profiles of the microbial biogeography of the gut have been limited by sample collection methods. The majority of human gut microbiome studies have been performed on whole shed stool or on samples collected during colonoscopy procedures. While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel. Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer (@Jalanka2014, @Harrell2012). As such, what little information we do have about the biogeographical distribution of the microbes in the proximal and distal colon is confounded by the bowel preparation procedure.

We combined these approaches to address the limitations of previous studies and effectively characterize the microbiome in the lumen and mucosa of the proximal and distal healthy human colon. Our design used unprepared colonoscopy techniques to sample each location of the gut without prior disruption of the native bacteria in 20 healthy volunteers. To address noise created by a diverse set of human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location. We found that our classification models were able to separate mucosal and lumenal samples as well as differentiate between sides of the colon based on populations of specific microbes. By identifying the specific microbes we are poised to ask if and how the presence or disruption of the microbes at each site contribute to the development of the specific tumor subtypes of CRC in the proximal and distal human colon.

@@

#### Microbial membership and diversity of the proximal and distal colon 
Proximal and distal lumenal and mucosal samples were collected from 20 healthy humans that had not undergone bowel preparation(Figure 1). Participants also collected stool at home one week prior to the procedure. To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample. The relative abundance of each phyla for each site is shown in Figure 2A. This data combines the samples from all participants in the study. Each site is primarily dominated by _Firmicutes_ and _Bacteriodetes_, consistent with known variability in human microbiome research (cite). Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B). That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa. Therefore we could not identify a clear pattern of changes in microbial diversity along the gut axis.

To compare similarity between sides (left or right) or sites (lumen or mucosa), we calculated theta YC distances from OTU abundances and compared distances for all individuals. Again, across all patient samples we observed a range of theta YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis. However, when comparing median distances between the right lumen and mucosa, the left versus right lumen, the left versus right mucosa and the left lumen and mucosa, we found that the right lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment).

Next, we calculated thetaYC distances to examine how each sample compared to the home-collected exit stool. Amidst variability between patients, we did identify significantly smaller thetaYc distance between the left lumenal sample and the exit stool (Figure 3B, P < 0.05, Wilcoxon, BH adjustment). Furthermore, there was an even larger difference in the comparisons of the left mucosa to the exit stool, indicating that the mucosa is different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment). To determine what factors may be driving the differences seen among the samples, we compared thetaYC distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal). We found that samples from one individual were overall much more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon (@Eckberg2005, @deCarcer2010, @Zhang2013). Thus interpersonal variation between subjects drives the differences between samples more than sample site or location. Overall, the results of the similarity analysis suggest that the contents of the left lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different.

#### Random Forest classification models identify important OTUs on each side
To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest models built on OTU abundances. We built the first model to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A). The constructed model used ((Xopt)) features for the right and ((Xopt)) for the left. The models performed well when classifying these samples (AUC = 0.797 and AUC = 0.980, respectively). The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the right lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B). The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4C). These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.

@@

#### Bacterial OTUs associated with cancer are found in healthy individuals 
Given that specific bacterial species have been associated with colorectal cancer, we probed our sample set for these OTUs. Among our 100 samples, the most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which aligns via BLASTn to _Fusobacterium nucleatum subsp animalis_. It was the only sequence to align with a _F. nucleatum_ strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors. (@Castellarin2011). The _Fusobacterium_ positive samples were located on the right and left mucosa and ranged up to 1% of the sample (Figure 6A). OTU00152 is _Porphyromonas_ and the most frequent sequence in that OTU aligns to _Porphyromonas asacharolytica_, another bacterium commonly detected and isolated from colorectal tumors. OTU152 was only detected on the left mucosa, and in fact was one of the OTUs the classification model identified as separating left and right sides (Figure 6B). Samples that were positive for _P. asacharolytica_ ranged in abundances from 0.01% - 16%. Thus, cancer-associated OTUs could be found in our sample set of 20 healthy individuals.

@@

When examining the relative abundance of the different phyla at each site, there was a wide amount of variation for each phyla with communities primarily dominated by the Bacteriodes and Firmicutes. This likely reflects not only the variability between human subjects, casued by differences in age, gender, diet, but also reports of microbial "patchiness" in the gut microbiome. Several previous studies have noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other (@Hong2011). Similar patchiness is also observed in lumenal contents and fecal samples themselves; there is observed separation of different interacting microbes along the length of a stool sample, for instance (@Stearns2011). That said, across our samples the mucosal samples harbor more Proteobacteria, consistent with previous studies comparing mucosal swabs to lumenal content in humans (@Albenberg2014). Hence, the conclusions we can draw from phyla analysis are likely confounded by differences in sampling and patchiness between subjects.

@@

We detected _F. nucleatum_ and _P. asacharolytica_ in 8 and 5 of our subjects, respectively. These bacteria have been shown to be predictive of colorectal cancer in humans (@Baxter2016) and have oncogenic properties in cell culture and in mice (@Sears2014). Interestingly, while _F. nucleatum_ was found on both sides of the colon, _P. asacharolytica_ was only detected in the left mucosa. Not much is known about the distribution of _P. asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous left mucosa (@Albenberg2014). In studies examining bacteria on colorectal cancer tumors, _F. nucleatum_ is more commonly detected on right-sided tumors, and distribution of _F. nucleatum_ decreases along the colon to rectum (@Mima2016).  Of the (8) (40%) individuals positive for _F. nucleatum_ in this study, the bacteria was spread across the right mucosa, left lumen and left mucosa. The presence of _F. nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers. Because of the spatial distribution of the _F. nucleatum_ in our sample set, we cannot develop a model for the role of _F. nucleatum_ in the healthy colon. Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found both on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis (@Dejea2014). Finally, _Fusobacteria_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model (@Whitmore2014, @Flynn2016). Thus, further analysis of the distribution and activities of these pathogens may elucidate a mechanism for development of CRC tumors in the proximal or distal colon.
